{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-12 thesis/quicksand.tex
copy in iterative deepening proof

@@

To introduce the Iterative Deepening algorithm,
I will first show a simple approach for handling new PPs in the absence of any CPU budget restriction.

Given unlimited CPU time for testing,
one would always want to switch to the new maximal state space whenever adding a new preemption point.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known PPs would be duplicate work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API PPs,
and invoked whenever a new data-race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
% TODO: clarify sec reference to -M flag
\cref{sec:quicksand-implementation} discusses this execution mode further.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all interleavings possible with its PPs were tested;
\cref{sec:quicksand-discussion} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

\subsection{Initial preemption points}

@@

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	\Input{$j_0$, the currently-running job}
	\Input{\AllJobs, the set of all existing (or completed) jobs}
	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
	\If{$\forall j \in \AllJobs,$
	\GetPPSet{$j_0$} $\cup$ $\alpha$
	$\not\subseteq$
	\GetPPSet{$j$}
	}{
		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
	}
	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
	}
	\caption{Adding new jobs with data-race PPs.}
	\label{alg:handledatarace}
\end{algorithm}

Furthermore,
Iterative Deepening allows not always strictly increasing the number of preemption points
whenever a new data race is identified.
For each instruction involved in a data race, Quicksand adds two new jobs:
a ``small'' job to preempt on that instruction only (line~5),
and a ``big'' job to preempt on that instruction as well as each preemption point used by the reporting job (line~2).
%
Hence,
each {\em pair} of racing accesses will spawn four new jobs.
\Cref{fig:new-dr-jobs} visualizes the resulting overall workflow in Quicksand,
including the four such jobs resulting from one data race report.%
\footnote{As an optimization,
though the big jobs should be expected to uncover more data races and in turn
produce even bigger jobs still,
small jobs should be forbidden from ``reproducing'',
as their purpose is only fast heuristic bug-finding rather than exhaustive coverage;
see {\tt handle\_data\_race()} in {\tt messaging.c}.}
%
The rationale of spawning multiple jobs is that one cannot know in advance which will be most fruitful:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original PPs were required to expose it.
In practice, I have observed many bugs found quickly by these small jobs,
and many other bugs missed by the small jobs found eventually by the big jobs.
This phenomenon motivates Iterative Deepening to prioritize jobs at run-time.

@@

Landslide classifies data-race candidates as {\em both-order} or {\em single-order},
as defined in prior work \cite{portend},
based on whether it observed the racing instructions ordered in both possible sequences or only one
in the original state space, respectively.
Single-order candidates are more likely to be false positives (\cref{sec:background-datarace}),
although preempting during the access itself is necessary to say for sure.
Hence, Quicskand add preemption points for both types of candidates,
and heuristically prioritizes jobs with both-order data races
over those with only single-order data races.
The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
For single-order races, Quicksand does not initially add a PP for the later access at all:
if preempting on the first access is capable of reordering the race,
it will be updated to both-order in the new state space, and the second preemption point will be added then.
\cref{sec:warpzone-heuristics} will discuss opportunities for future work to expand
these heuristics with more nuanced search strategies still.

@@

Adding new data-race preemption points in a feedback loop can uncover bugs
not previously reachable by preempting on synchronization APIs alone,
as some prior model checkers do \cite{dbug-ssv},
but how does it compare to the other extreme end of the trade-off,
that is,
committing in advance to preempt on every single shared memory access \cite{inspect}?
It turns out,
assuming sufficient CPU budget,
Iterative Deepening can in principle expose every possible program behaviour that
even that latter approach can find,
providing an equally strong verification guarantee.
This section presents a proof of this claim (\cref{sec:quicksand-convergence}),
as well as a supplementary proof (\cref{sec:quicksand-realloc})
of the soundness of pruning reallocation false positives discussed previously (\cref{sec:quicksand-id-realloc}).

@@

\subsection{Convergence}
\label{sec:quicksand-convergence}

\subsubsection{Equivalence}

% TODO: decide if this shouldn't be reordered befor econvergence?
\subsubsection{Saturation}

\subsection{Reallocation false positives}
\label{sec:quicksand-realloc}

>>>>

% TODO: use this somehow
Committing in advance to
preempt on every instruction is certain to expose any possible bug under the given test case,
but invites massive state space explosion.
Even as DPOR helps to skip equivalent interleavings of non-conflicting transitions,
DPOR itself is $O(n^2)$ in the number of preemption points in a single execution,
which is not compatible with such an approach.
Accordingly, stateless model checkers must find more efficient ways to be able to uncover bugs such as these.

@@

To introduce the Iterative Deepening algorithm,
I will first show a simple approach for handling new preemption points in the absence of any CPU budget restriction.

Given unlimited CPU time for testing,
one would always want to switch to the new maximal state space whenever adding a new preemption point.
The maximal state space is guaranteed to subsume all execution sequences reachable in any subset state space,
so considering any incomplete subset of the known preemption points would be duplicate work.
\Cref{alg:algorithm0} demonstrates this na\"ive approach.
It is seeded with the set of all statically-known synchronization API preemption points,
and invoked whenever a new data-race candidate is found.
%
The upcoming proofs in \cref{sec:quicksand-soundness},
being concerned with the verification guarantee provided when the search may complete within the CPU budget,
are based on this simple version of Iterative Deepening.
The user may also wish to configure her testing tool to prefer this approach, at her discretion,
such as when she believes all bugs have been fixed and wants a verification as fast as possible;
% TODO: clarify sec reference to -M flag
\cref{sec:quicksand-implementation} discusses this execution mode further.

@@

However, in many tests of even modestly-sized programs,
full verification is not feasible,
and focusing on the maximal state space alone is likely to be fruitless.
%
Hence, Iterative Deepening also allows for prioritizing subset jobs
based on number of preemption points, ETA, and whether data race candidates are included among their preemption points.
It relies on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing a large fraction of interleavings for each.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only her total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.
Quicksand seeks to maximize completed state spaces,
as each one serves as a guarantee that all possible interleavings therein were tested;
\cref{sec:quicksand-discussion} discusses some limitations of this approach.
The next three subsections will show how to schedule these smaller jobs
based on their preemption points and ETAs.

@@

\subsection{Initial preemption points}
\label{sec:quicksand-initial-pps}

@@

\newcommand\AllJobs{\ensuremath{\mathcal{J}}}
\begin{algorithm}[t]
	\SetKwInOut{Input}{Input}
	\Input{$j_0$, the currently-running job}
	\Input{\AllJobs, the set of all existing (or completed) jobs}
	\Input{$\alpha$, an instruction reported by the MC as part of a racing access pair}
	\If{$\forall j \in \AllJobs,$
	\GetPPSet{$j_0$} $\cup$ $\alpha$
	$\not\subseteq$
	\GetPPSet{$j$}
	}{
		AddNewJob(\GetPPSet{$j_0$} $\cup$ $\alpha$, HeuristicPriority($\alpha$)) \\
	}
	\If{$\forall j \in \AllJobs,$ \GetPPSet{$j$} $\neq \{yield, \alpha\}$}{
		AddNewJob($\{yield, \alpha\}$, HeuristicPriority($\alpha$))
	}
	\caption{Adding new jobs with data-race preemption points.}
	\label{alg:handledatarace}
\end{algorithm}

Furthermore,
Iterative Deepening allows not always strictly increasing the number of preemption points
whenever a new data race is identified.
For each instruction involved in a data race, Quicksand adds two new jobs:
a ``small'' job to preempt on that instruction only (line~5),
and a ``big'' job to preempt on that instruction as well as each preemption point used by the reporting job (line~2).
%
Hence,
each {\em pair} of racing accesses will spawn four new jobs.
\Cref{fig:new-dr-jobs} visualizes the resulting overall workflow in Quicksand,
including the four such jobs resulting from one data race report.%
\footnote{As an optimization,
though the big jobs should be expected to uncover more data races and in turn
produce even bigger jobs still,
small jobs should be forbidden from ``reproducing'',
as their purpose is only fast heuristic bug-finding rather than exhaustive coverage;
see {\tt handle\_data\_race()} in {\tt messaging.c}.}
%
The rationale of spawning multiple jobs is that one cannot know in advance which will be most fruitful:
while the big job risks not completing in time,
the small job risks missing the data race entirely if the original preemption points were required to expose it.
In practice, I have observed many bugs found quickly by these small jobs,
and many other bugs missed by the small jobs found eventually by the big jobs.
This phenomenon motivates Iterative Deepening to prioritize jobs at run-time.

@@

Landslide classifies data-race candidates as {\em both-order} or {\em single-order},
as defined in prior work \cite{portend},
based on whether it observed the racing instructions ordered in both possible sequences or only one
in the original state space, respectively.
Single-order candidates are more likely to be false positives (\cref{sec:background-datarace}),
although preempting during the access itself is necessary to say for sure.
Hence, Quicskand add preemption points for both types of candidates,
and heuristically prioritizes jobs with both-order data races
over those with only single-order data races.
The HeuristicPriority($\alpha$) call in \Cref{alg:handledatarace} corresponds to this strategy.
For single-order races, Quicksand does not initially add a preemption point for the later access at all:
if preempting on the first access is capable of reordering the race,
it will be updated to both-order in the new state space, and the second preemption point will be added then.
\cref{sec:warpzone-heuristics} will discuss opportunities for future work to expand
these heuristics with more nuanced search strategies still.

@@

Adding new data-race preemption points in a feedback loop can uncover bugs
not previously reachable by preempting on synchronization APIs alone,
as some prior model checkers do \cite{dbug-ssv},
but how does it compare to the other extreme end of the trade-off,
that is,
committing in advance to preempt on every single shared memory access \cite{spin,inspect}?
It turns out,
assuming sufficient CPU budget,
Iterative Deepening can in principle expose every possible program behaviour that
even that latter approach can find,
providing an equally strong verification guarantee.
This section presents a proof of this claim (\cref{sec:quicksand-convergence}),
as well as a supplementary proof (\cref{sec:quicksand-realloc})
of the soundness of pruning reallocation false positives discussed previously (\cref{sec:quicksand-id-realloc}).

@@

\subsection{Convergence to total verification}
\label{sec:quicksand-convergence}

The proof of Iterative Deepening's soundness is in two parts.
In the first part, I prove that for any possible interleaving
one could execute with preemptions anywhere,
an equivalent interleaving must exist using only data-race and synchronization preemption points.
In the second, I prove that starting from synchronization preemption points only,
Iterative Deepening must eventually reach a state space containing such an interleaving,
no matter how many data races are involved.

\newcommand\ppnext[1]{\ensuremath{\mathsf{next}(#1)}}
\newcommand\ppinstr[1]{\ensuremath{\mathsf{instr}(#1)}}
\newcommand\ppothers[1]{\ensuremath{\mathsf{others}(#1)}}

Given a preemption point $p$,
let $\ppnext{p}$ denote the next transition after $p$ executed by the thread which ran immediately before $p$,
let $\ppinstr{p}$ denote the first instruction of $\ppnext{p}$,
and let $\ppothers{p}$ denote the transitions by other threads between $p$ and $\ppnext{p}$.

% TODO: decide if these subsection headers are good, or, not
\subsubsection{Equivalence}

\begin{lemma}[Equivalence of non-data-race preemption points]
	For any thread interleaving possible by preempting on any instruction,
	there exists an equivalent interleaving which uses only data-race and synchronization API preemption points.
        \label{lem:relevant}
\end{lemma}

\begin{proof}
Let $p$ be the first preemption point in the given interleaving
such that $\ppinstr{p}$ is not a data race with $\ppothers{p}$ nor is a synchronization API boundary.
Because $\ppinstr{p}$ is not a synchronization boundary,
no lock can be held during $\ppothers{p}$ that was also held by the first thread across $p$.
Hence, because $\ppinstr{p}$ is not a data race, it cannot be a shared memory conflict with $\ppothers{p}$ at all.
Let $i$ be the first instruction among $\ppnext{p}$ which is such a conflict, or a synchronization boundary.
If $i$ is a shared memory conflict, it must be a data race, for the same reasoning as above.
Modify the input interleaving by reordering $\ppinstr{p}$ until $i$, not including $i$, to before $\ppothers{p}$.
% TODO: fix this handwavey citation to at least point to something that says what soundness of dpor means
By the soundness of DPOR \cite{dpor}, this is equivalent to the input interleaving.
In other words, $p$ has been transformed into $p'$ such that $\ppnext{p'} = i$,
which is a data race or synchronization boundary.
All other preemption points in the input trace can be inductively converted in the same manner.
\end{proof}

\subsubsection{Saturation}

For Iterative Deepening to ``eventually'' reach a certain state space,
all data-race preemption points involved must be {\em reachable} during the test.

\begin{definition}[Reachability]
	A data race candidate, and its associated preemption point(s),
	are reachable if it will be identified by a model checker
	configured to preempt only on already-reachable preemption points.
\end{definition}

Initially, the statically-available synchronization API preemption points (\cref{sec:quicksand-initial-pps})
are reachable.
Reachability of data-race preemption points is transitive.

\begin{lemma}[Saturation of data races]
        Given any interleaving comprising only data-race and synchronization API preemption points,
        all involved preemption points are reachable.
        \label{lem:saturation}
\end{lemma}

\begin{proof}
Induct on the preemption points according to the order of their preemptions during an execution sequence.
Given that the interleaving prefix preceding some point $p$ is reachable,
either $p$ must be reachable,
or a new data race among $\ppothers{p}$ will be newly reachable. %, not previously reachable, is now reachable.
The latter condition suffices because in a finitely-sized codebase,
there must be finitely many unique racing instruction pairs.
%, so induction on the number of new data-race preemption points among $\ppothers{p}$ will make $p$ itself reachable.

First, $p$ must be ``coalesced'' away, as well as any other not-yet-reachable points in $\ppothers{p}$.
Consider the alternate interleaving in which the first thread executes past $p$ until the first already-reachable point,
then the other threads among $\ppothers{p}$ execute the same way.
This interleaving's preemption points are all reachable,
so a state space $\mathcal{S}$ containing it will be tested.

If $p$ is a not-yet-reachable data-race preemption point,
it must be possible for some other thread to execute a data-racing instruction with $\ppinstr{p}$.
If this conflict was observed in the state space containing our coalesced interleaving, $p$ is reached.
Otherwise, appeal to the soundness property of DPOR:
%(\cref{sec:landslide-dpor}): % FIXME ugh, this section reference is so handwavy/imprecise
If a program behaviour is possible by interleaving threads at the boundaries of the given transitions,
it will be tested in the containing state space.
By contrapositive, to expose this behaviour,
one or more preemptions must occur in the middle of some transition, rather than at the boundaries.

Let us now see by contradiction that there cannot be {\em multiple} data-race preemption points
which must all be enabled before either data race can be identified;
i.e., a circular dependency.
Assume there does not exist a single transition $t_1 \in \mathcal{S}$
which alone can be split into $\{t_1',t_1''\}$ by a point $q$,
such that another thread's concurrent transition $t_2$ conflicts with $t_1''$.
By the soundness of DPOR, because all $t_2$s are independent with $t_1''$, $\mathcal{S} \equiv \mathcal{S} \cup q$.
Replacing $\mathcal{S}$ with $\mathcal{S} \cup q$ in the above assumption shows that no {\em pair} of new $q$s would expose new program behaviour, and inductively, no set of $q$s of any size, which contradicts the previous paragraph.

Hence, a single new not-yet-reachable data race is reachable in $\mathcal{S}$. Hence $p$ will be reached.
\end{proof}

\subsubsection{Convergence}

I name the overall soundness property ``convergence''
in reference to the way it must eventually arrive,
after potentially many iterated state spaces,
at full verification strength.

\begin{theorem}[Convergence]
	If a bug can be exposed by any thread interleaving possible
	by preempting on all instructions during a specific test,
	Iterative Deepening will eventually test an equivalent interleaving which exposes the same bug.
        \label{thm:convergence}
\end{theorem}

\begin{proof}
	For any possible interleaving,
	Lemma \ref{lem:relevant} provides an equivalent one with only data-race and synchronization preemption points,
	and Lemma \ref{lem:saturation} proves all involved preemption points are reachable.
	Hence, Iterative Deepening will eventually test a state space containing the equivalent buggy interleaving.
\end{proof}

And thus Iterative Deepening is sound. % :)

\subsection{Suppressing reallocation false positives}
\label{sec:quicksand-realloc}
